---
title: "Final project: Does benthic diatom diversity respond to expanding littoral habitat?"
author: "Vaclava Hazukova"
date: "2022-12-16"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Lakes in West Greenland have experienced abrupt environmental shifts as a result of rapid temperature increase during the 1990s. Among the non-linear, ecological shifts are prolonged open-water period, higher lakewater temperatures, and increasing water clarity due to decline in dissolved organic carbon concentrations. Higher transparency of water also leads to expansion of the littoral area, illuminated shoreline habitat that benthic diatoms use. 

We collected two sets of surface sediments from 13 lakes across the Kangerlussuaq region in West Greenland; one set was collected in 1996 and characterizes period prior to rapid climate shift, the second set was collected in 2013 and represents period after the rapid step-shift occurred. We analyzed diatoms in each set, distinguishing between planktic and benthic species. Considering that surface sediments represent accumulation over approximately 10 years prior collection in these lakes, the diatom assemblages are inter- and intra-annually integrated. 

Paired with the surface sediments, we also have limnological data; isolated measurements from the late 90s, and continuous monitoring starting in 2013. Using dissolved organic carbon concentration data, we are able to calculate shifts in light penetration in each lake. Conservatively, we use 10% PAR depth to represent benthic diatom habitat. Using this information, we are able to calculate how lake area available to benthic diatoms changed between the two periods. 

The expectation is that as habitat expands, benthic diatom species richness will increase. To test the idea, we will use accumulated species curves and compare the number of species predicted by the curve based on diatom data from 1996 with data from 2013. 

```{r libraries and read in data}
library(tidyverse); library(lubridate)

# surface sdiment diatom data

surfsed <- 
  read_csv('data/processed/surface_sediments_combined.csv')

# environmental data

# a) water chemistry (layer-level data)

env <- 
  read_csv('data/raw/Greenland_2013-2019_env_MASTER.csv') %>% 
  mutate(year = year(Date)) %>% 
  filter(year == 2013 & Layer == 'epi') %>% 
  select(-ice, -iceout) %>% 
  rename(lake = Lake) %>% 
  group_by(lake) %>% 
  summarise(across(c(chl, DOC, TP, NO3, PAR), mean, na.rm = T))

# b) DOC decline

doc_change <- 
  read_csv('data/processed/doc_calculated.csv') %>% 
  select(c(1:4, 26:28))


# combine bathymetries and doc change for the 13 lakes where data are available

bathy <- 
  read_csv("data/raw/bathy.csv") %>% 
  inner_join(doc_change %>% select(lake, par10_96, par10_13),
            by = 'lake')

# diatom traits

trait <- read_csv('data/raw/diatom_traits_analysis.csv')

```

```{r}

# calculate volume of lake with > 10% PAR

PAR_area <- map_dfr(
  unique(bathy$lake),
  function(x) {
    temp <- 
      bathy %>% 
      filter(lake == x)
    
    a1 = temp %>% filter(upper == 1 ) %>% pull(area)
    
    #1996
    depth_top = temp %>% distinct(trunc(par10_96, 0)) %>% pull()
    depth_bottom = depth_top + 1 
    a_par_round = temp %>% filter(upper == depth_top) %>% pull(area)
    extra = temp %>% summarise(par10_96 - depth_top) %>% distinct() %>% pull()
    diff = temp %>% summarise(area = area[upper == depth_top] - area[upper == depth_bottom]) %>% pull() 
    a_total_at_par = a_par_round - diff*extra
    a_par_96 = a1-a_total_at_par
    
    #2013
    depth_top_13 = temp %>% distinct(trunc(par10_13, 0)) %>% pull()
    depth_bottom_13 = depth_top_13 + 1 
    a_par_round_13 = temp %>% filter(upper == depth_top_13) %>% pull(area)
    extra_13 = temp %>% summarise(par10_13 - depth_top_13) %>% distinct() %>% pull()
    diff_13 = temp %>% summarise(area = area[upper == depth_top_13] - area[upper == depth_bottom_13]) %>% pull() 
    a_total_at_par_13 = a_par_round_13 - diff_13*extra_13
    a_par_13 = a1-a_total_at_par_13
    
    tibble(
      lake = unique(temp$lake),
      PAR_A_1996 = a_par_96,
      PAR_A_2013 = a_par_13
      )
    
  }) %>% 
  mutate(diff_A = PAR_A_2013 - PAR_A_1996,
         diff_A_percent = ((PAR_A_2013 - PAR_A_1996)/PAR_A_2013*100)) %>% 
  left_join(doc_change %>% select(lake, par10_96, par10_13),
            by = 'lake') %>% 
  mutate(diff_par_depth = ((par10_13-par10_96)/par10_13*100)) %>% 
  rename(par10_1996 = par10_96, par10_2013 = par10_13)
  
# join with lake characteristics 

id <- 
  read_csv("data/raw/lake_id.csv") %>% 
  mutate(GR = A^0.25/maxZ,
         CEI = A/avgZ) %>% 
  rename(lake = Lake) %>% 
  left_join(PAR_area,
            by = 'lake')

```


```{r}

# assign diatom status based on presence or ansence in the early 96 dataset

status_diatom <- 
  surfsed %>%
  pivot_wider(names_from = year,
              values_from = freq,
              names_prefix = 'year_') %>% 
  filter(!year_2013 == 0 | !year_1996 == 0) %>% 
  mutate(status = case_when(
    year_2013 > 0 & year_1996 == 0 ~ 'immigration',
    year_2013 == 0 & year_1996 > 0 ~ 'extinction',
    year_2013 > 0 & year_1996 > 0 ~ 'stay'))

# total number of species found in each year and for each lake 

numbers_of_species_tot <-
  surfsed %>% 
  filter(!freq == 0) %>% 
  group_by(lake, year) %>%
  summarise(tot_n = n())

# grouped als oby status 

numbers_of_species <-
  surfsed %>% 
  filter(!freq == 0) %>% 
  left_join(status_diatom %>% select(-year_2013, -year_1996),
            by = c('taxon', 'lake')) %>% 
  group_by(lake, year, status) %>%
  summarise(count = n())

# show count
surfsed %>% 
  filter(!freq == 0) %>% 
  left_join(status_diatom %>% select(-year_2013, -year_1996),
            by = c('taxon', 'lake')) %>% 
  group_by(year, status) %>%
  distinct(taxon) %>% 
  summarise(count = n())

# create the final joint dataset; assign species tratis, filter benthics with more than 0 occurrences

benthos <- 
  surfsed %>%
  left_join(trait %>% select(taxon, pl, pl2, biovol, 10:28),
            by = 'taxon') %>% 
  filter(pl2 == 0) %>% 
  group_by(lake, year) %>% 
  filter(freq > 0) %>% 
  summarise(benthic_count = n()) %>% 
  pivot_wider(names_from = year,
              values_from = benthic_count,
              names_prefix = 'year_') %>% 
  mutate(diff_benthic_diatoms_percent = (year_2013 - year_1996)/year_2013 * 100,
         diff_benthic_diatoms = year_2013 - year_1996) %>% 
  inner_join(PAR_area,
             by = 'lake') %>% 
  inner_join(numbers_of_species %>% 
               filter(status %in% c('immigration')),
             by = 'lake') %>% 
  inner_join(numbers_of_species_tot %>% 
               filter(year == 2013) %>% select(-year),
             by = 'lake')

# SAD 
surfsed %>% 
  group_by(year) %>% 
  ggplot(aes(x = freq)) +
  geom_histogram() +
  facet_wrap(~year)


```


```{r }
benthos %>% 
  ggplot(aes(x = diff_A, y = count)) +
  geom_point() + 
  geom_smooth(method = lm) +
  geom_abline(slope = 1, intercept = 0) +
  ggrepel::geom_text_repel(aes(label = lake)) +
  labs(y = 'Number of new species (2013 - 1996)', x = 'change in 10% PAR surface area (m2)') +
  theme_bw()

```
The first plot shows the number of 'new' species (immigration) that appeared in each lake in 2013 relative to 1996 against the change in habitat. Very, very scattered positive relationship. 

```{r}
surfsed %>%
  left_join(trait %>% select(taxon, pl, pl2, biovol, 10:28),
            by = 'taxon') %>% 
  filter(freq > 0 &
           !is.na(pl)) %>% 
  group_by(lake, year) %>% 
  summarise(pb = sum(freq[pl == "1"]) / sum(freq[pl == "0"])) %>% 
  inner_join(PAR_area %>% 
               select(c(1:3)) %>%
               pivot_longer(cols = 2:3,
                            names_to = 'year',
                            names_prefix = 'PAR_A_',
                            names_transform = as.double,
                            values_to = 'area'),
             by = c('lake','year')) %>% 
  ggplot(aes(x = area, y = pb, group = lake)) +
  geom_point(aes(color = factor(year)), size = 3) +
  scale_color_manual('Year', values = c('skyblue3', 'orangered')) + 
  geom_line(aes(group = lake)) +
  labs(y = 'P:B ratio', x = '10% PAR surface area (m2)') +
  theme_bw()

```

In the second plot, we look at shifts in plankton vs. benthic species ratio (the expectation being that as habitat for benthic increase, we will see a decrease in this metric), and it's all over the place. Some lakes seem to have a very similar direction and magnitude of shifts; however, in others, planktic ratio increases or does not change. 

Species accumulation curves: we are resampling how species richness increases with each additional lake and comparing the two datasets: 1996 and 2013. 

```{r rarefy }

# rarefy ------------------------------------------------------------------

surf_par <- 
  surfsed %>%
  left_join(trait %>% select(taxon, pl, pl2, biovol),
            by = 'taxon') %>% 
  filter(pl2 == 0) %>% 
  filter(freq > 0) %>% 
  inner_join(PAR_area %>% 
               select(c(1:3)) %>%
               pivot_longer(cols = 2:3,
                            names_to = 'year',
                            names_prefix = 'PAR_A_',
                            names_transform = as.double,
                            values_to = 'area'),
             by = c('lake','year'))

surf_par_96 <- 
  surf_par %>% 
  filter(year == 1996)

cum_sp <- 
  parallel::mclapply(1:100, mc.cores = 8, FUN = function(j) {
    
    names_lake <-  sample(unique(surf_par_96$lake))
    
    x <- 
      lapply(1:length(names_lake), function(i) {
        temp_name <- names_lake[1:i]
        A = surf_par_96 %>% filter(lake %in% temp_name) %>% distinct(lake,area) %>% summarise(area = sum(area)) %>% pull()
        S = surf_par_96 %>% filter(lake %in% temp_name & freq > 0) %>% distinct(taxon) %>% summarise(n = n()) %>% pull()
        run = j
        c(A,S,run)
      }) 
    do.call(rbind, x) })

cum_sp <- do.call(rbind, cum_sp)

surf_par_13 <- 
  surf_par %>% 
  filter(year == 2013)

cum_sp_13 <- 
  parallel::mclapply(1:100, mc.cores = 8, FUN = function(j) {
    
    names_lake <-  sample(unique(surf_par_13$lake))
    
    x <- 
      lapply(1:length(names_lake), function(i) {
        temp_name <- names_lake[1:i]
        A = surf_par_13 %>% filter(lake %in% temp_name) %>% distinct(lake,area) %>% summarise(area = sum(area)) %>% pull()
        S = surf_par_13 %>% filter(lake %in% temp_name & freq > 0) %>% distinct(taxon) %>% summarise(n = n()) %>% pull()
        run = j
        c(A,S,run)
      }) 
    do.call(rbind, x) })

cum_sp_13 <- do.call(rbind, cum_sp_13)

cum_sp <- cum_sp %>% 
  as_tibble() %>% 
  mutate(year = '1996')

comb <- cum_sp_13 %>% 
  as_tibble() %>% 
  mutate(year = '2013') %>% 
  bind_rows(cum_sp) %>% 
  rename(area = V1, S = V2, run = V3) %>% 
  mutate(code = str_c(year, run, sep = ''))
   

comb %>% 
  ggplot(aes(x = log10(area), y = log10(S), group = code)) +
  geom_point(aes(color = year), size = 0.1)+
  geom_line(aes(group = code, color = year), linewidth = 0.1) +
  geom_smooth(aes(group = year, color = year), method = 'lm') +
  scale_color_manual('Year', values = c('skyblue1', 'orangered')) + 
  labs(y = 'log(S)', x = 'log(Cumulative area (m2))') +
  theme_bw()

```

```{r predict }

comb_log10_96 <- 
  comb %>% 
  filter(year == 1996) %>% 
  transmute(A_log10 = log10(area),
            S_log10 = log10(S))

comb_log10_13 <- 
  comb %>% 
  filter(year == 2013) %>% 
  transmute(A_log10 = log10(area),
            S_log10 = log10(S))

# (I shold probably use some sort of mixed model and use 'run' as a randon factor...)

lin_mod_96 <- lm(S_log10 ~ A_log10, data = comb_log10_96)

lin_mod_13 <- lm(S_log10 ~ A_log10, data = comb_log10_13)

pred_data <- 
  data.frame(A_log10 = seq(from = 4, 
                        to = 6.15, 
                        by = 0.05))

pred_data$pred_logS_96 <- predict(object = lin_mod_96, 
                               newdata = pred_data)

pred_data$pred_logS_13 <- predict(object = lin_mod_13, 
                               newdata = pred_data)

ggplot() +
  geom_point(data = comb, aes(x = log10(area), y = log10(S), group = code, color = year), size = 0.1)+
  geom_line(data = comb, aes(x = log10(area), y = log10(S), group = code, color = year), linewidth = 0.1) +
 # geom_smooth(data = comb, aes(x = log10(area), y = log10(S), color = year), method = 'lm') +
  geom_smooth(data = pred_data, aes(x = A_log10, y = pred_logS_96), color = 'skyblue1', method = 'lm') +
   geom_smooth(data = pred_data, aes(x = A_log10, y = pred_logS_13), color = 'orangered', method = 'lm') +
  scale_color_manual('Year', values = c('skyblue1', 'orangered')) + 
  labs(y = 'log(S)', x = 'log(Cumulative area (m2))') +
  theme_bw()


```
The extended blue lines is basically touching the prediction line based on data from 2013; this suggests that habitat expansion is not an impossible explanation for the changes we see in species richness at the level of regional metacommunity. However, as seen above - within lake relationships do not seem to follow the same rules. 

Lastly, we can use maximum entropy theory of ecology to predict patterns of abundance in each dataset:

```{r}
library(meteR)

# 1996

diatom_mete <- 
  surfsed %>% 
  inner_join(id %>% 
               filter(!is.na(par10_1996)) %>% 
               select(lake, lat, long, A, PAR_A_1996, PAR_A_2013, diff_A),
             by = 'lake') %>% 
  left_join(trait %>% 
              select(taxon, pl, biovol),
            by = 'taxon') %>% 
  filter(!is.na(biovol) &
           year == 1996)


esf1 <- meteESF(spp = diatom_mete$taxon,
                abund = diatom_mete$freq,
                power = diatom_mete$biovol^(3/4),
                minE = min(diatom_mete$biovol^(3/4)))
esf1  

sad1 <- sad(esf1)
sad1

plot(sad1, ptype = 'rad', log = 'y')
plot(sad1, ptype = 'cdf', log = 'x')
  
# 2013

diatom_mete <- 
  surfsed %>% 
  inner_join(id %>% 
               filter(!is.na(par10_1996)) %>% 
               select(lake, lat, long, A, PAR_A_1996, PAR_A_2013, diff_A),
             by = 'lake') %>% 
  left_join(trait %>% 
              select(taxon, pl, biovol),
            by = 'taxon') %>% 
  filter(!is.na(biovol) &
           year == 2013)



esf1 <- meteESF(spp = diatom_mete$taxon,
                abund = diatom_mete$freq,
                power = diatom_mete$biovol^(3/4),
                minE = min(diatom_mete$biovol^(3/4)))
esf1  

sad1 <- sad(esf1)
sad1

plot(sad1, ptype = 'rad', log = 'y')
plot(sad1, ptype = 'cdf', log = 'x')

```
Interestingly, the 2013 data are predicted by METE much better that the 'baseline' dataset from 1996 that has more rare species than predicted! Why is that? 

## Next steps: 
1) Need to think about the differing predictions by METE for the two datasets? Why is that earlier dataset does not fit predictions?
2) Look at beta diversity patterns - which species persist? which species get replaced between the two time periods?
3) Gather more limnological data and refine the calculations changing habitat extent (now, we assume that the quality of DOC does not change, but that is likely a false + changing quality will have a large impact on the color of DOC!)
